pyFAST: A Modular and Efficient Python Library for Time Series Analysis

### 2025-03-12 paper/paper.tex
update

\begin{abstract}
Modern time series analysis demands flexible, efficient, and extensible software, yet existing Python libraries often fall short, lacking modularity and support for cutting-edge models, especially for tasks like univariate forecasting and sparse data handling. To bridge this gap, we introduce \texttt{pyFAST}, a novel, research-driven Python framework that redefines modularity and efficiency in time series analysis.  \texttt{pyFAST} pioneers the integration of Large Language Model adaptations for univariate forecasting and offers native support for sparse data, alongside a comprehensive suite of state-of-the-art models, including Transformers and Graph Neural Networks.  Its unique modular architecture empowers users to seamlessly customize and extend the library, fostering rapid experimentation and methodological innovation. Benchmarking demonstrates \texttt{pyFAST}'s superior computational efficiency and competitive accuracy against leading libraries. Implemented in PyTorch and released under the MIT license, \texttt{pyFAST} is publicly available on GitHub with comprehensive documentation and examples, aiming to catalyze community contributions and accelerate advancements in time series research and applications.
\end{abstract}

@@

\section{Software Overview}
\texttt{pyFAST} is architected around modularity, with each component engineered to address a specific facet of time series analysis. This modular design philosophy enhances flexibility, maintainability, and extensibility, allowing users to tailor the library to diverse analytical tasks. The library’s core functionalities are organized into five modules, each designed with specific capabilities to ensure a cohesive and versatile framework. The \texttt{data} module is dedicated to data handling and preprocessing, featuring dataset classes tailored for STS, MTM, BDP, and STM scenarios, alongside a suite of scaling methods. This design allows for seamless integration of custom data pipelines.  The \texttt{model} module hosts a diverse collection of time series models, including classical, deep learning (CNNs, RNNs, Transformers), and GNN-based architectures for both UTS and MTS data. The \texttt{base} submodule provides fundamental building blocks for custom model creation.  The \texttt{train} module provides the \texttt{Trainer} class, streamlining model training with functionalities for validation, early stopping, checkpointing, and support for various optimization and learning rate scheduling techniques. The \texttt{metric} module offers a comprehensive suite of evaluation metrics for time series tasks, including specialized metrics for masked data, with the \texttt{Evaluator} class simplifying results reporting.  Lastly, the \texttt{visualize} module provides tools for visualizing time series data and model predictions, aiding in model analysis and interpretation. \texttt{pyFAST}'s modular architecture empowers researchers and practitioners to extend the library, customize modules, combine modules, and reproduce state-of-the-art models.

\section{Related Work and Comparison}

The landscape of time series modeling libraries and frameworks is rich and diverse, with several excellent tools available to researchers and practitioners. TensorFlow Time Series \citep{abadi2016tensorflow} provides a comprehensive and mature toolkit within the TensorFlow ecosystem, with a strong focus on forecasting and anomaly detection.  GluonTS \citep{alexandrov2020gluonts}, developed by Amazon, is a widely adopted library renowned for its extensive suite of probabilistic forecasting models and specialized tools for handling probabilistic time series prediction.  PyTorch Forecasting \citep{paszke2019pytorch} offers a collection of time series-specific layers, models, and training utilities within the PyTorch framework, emphasizing explainable and interpretable forecasting methodologies.  `sktime' \citep{loning2019sktime} stands out as a comprehensive Python toolbox encompassing a wide range of time series analysis tasks, including classical models, machine learning models, and sophisticated evaluation tools, with a strong emphasis on seamless integration with the scikit-learn ecosystem.  `tslearn' \citep{tavenard2020tslearn} specializes in time series-specific machine learning algorithms, offering tools for time series clustering, classification, representation learning, and related tasks.  `StatsForecast` \citep{garza2022statsforecast} and `NeuralForecast` \citep{olivares2022library_neuralforecast} are highly optimized libraries specifically engineered for fast and scalable forecasting, providing efficient implementations of both statistical and neural forecasting models, respectively, with a focus on computational performance.

@@

\section{Conclusion}
In summary, \texttt{pyFAST} is presented as a novel, modular, efficient, and comprehensive Python library, poised to significantly benefit both researchers and practitioners in the field of time series analysis. Its core strengths lie in its research-driven flexibility, diverse model library, and ease of use.  Our benchmarking results confirm its practical utility and competitive efficiency against state-of-the-art baselines. We are confident that \texttt{pyFAST} will serve as a valuable, high-performance, and adaptable platform, fostering future innovations in time series research and a wide range of applications. \texttt{pyFAST}, implemented in PyTorch and MIT-licensed, is publicly accessible on GitHub at \href{https://github.com/anonymous-code-for-peer-review/pyFAST}{GitHub}, accompanied by extensive resources to support and expand its community of users and contributors.

>>>>

\begin{abstract}
Modern time series analysis demands flexible, efficient, and extensible software, yet existing Python libraries often fall short, lacking modularity and support for cutting-edge models, especially for tasks like univariate forecasting and sparse data handling. To bridge this gap, we introduce \texttt{pyFAST}, a novel, research-driven Python framework that redefines modularity and efficiency in time series analysis.  \texttt{pyFAST} pioneers the integration of Large Language Model adaptations for univariate forecasting and offers native support for sparse data, alongside a comprehensive suite of state-of-the-art models, including Transformers and Graph Neural Networks.  Its unique modular architecture empowers users to seamlessly customize and extend the library, fostering rapid experimentation and methodological innovation. Benchmarking demonstrates \texttt{pyFAST}'s superior computational efficiency and competitive accuracy against leading libraries. \texttt{pyFAST} is publicly available on GitHub at \href{https://github.com/anonymous-code-for-peer-review/pyFAST}{GitHub}, implemented in PyTorch and released under the MIT license, with comprehensive documentation and examples, aiming to catalyze community contributions and accelerate advancements in time series research and applications.
\end{abstract}

@@

\section{Software Overview}
\begin{figure}[htbp]
    \centering
    \begin{mermaid}
graph LR
    subgraph pyFAST Library
        A[Data Module] --> A1[sts\_dataset.py]
        A --> A2[mtm\_dataset.py]
        A --> A3[scale.py]
        A --> A4[patch.py]
        B[Model Module] --> B1[uts/]
        B --> B2[mts/] --> B2a[transformer/]
        B2 --> B2b[gnn/]
        B --> B3[base/]
        C[Train Module] --> C1[Trainer (train.py)]
        C --> C2[EarlyStop (train.py)]
        D[Metric Module] --> D1[metric.py]
        D --> D2[mask\_metric.py]
        D --> D3[evaluate.py]
        E[Visualize Module] --> E1[plot\_in\_line\_chart (visualize.py)]
        E --> E2[plot\_comparable\_line\_charts (visualize.py)]
        F[Generative Module] --> F1[tvae.py]
        F --> F2[transvae.py]
    end
    style A fill:#f9f,stroke:#333,stroke-width:2px
    style B fill:#f9f,stroke:#333,stroke-width:2px
    style C fill:#f9f,stroke:#333,stroke-width:2px
    style D fill:#f9f,stroke:#333,stroke-width:2px
    style E fill:#f9f,stroke:#333,stroke-width:2px
    style F fill:#f9f,stroke:#333,stroke-width:2px
    style A1 fill:#ccf,stroke:#333,stroke-width:2px
    style A2 fill:#ccf,stroke:#333,stroke-width:2px
    style A3 fill:#ccf,stroke:#333,stroke-width:2px
    style A4 fill:#ccf,stroke:#333,stroke-width:2px
    style B1 fill:#ccf,stroke:#333,stroke-width:2px
    style B2 fill:#ccf,stroke:#333,stroke-width:2px
    style B3 fill:#ccf,stroke:#333,stroke-width:2px
    style B2a fill:#ddf,stroke:#333,stroke-width:2px
    style B2b fill:#ddf,stroke:#333,stroke-width:2px
    style C1 fill:#ccf,stroke:#333,stroke-width:2px
    style C2 fill:#ccf,stroke:#333,stroke-width:2px
    style D1 fill:#ccf,stroke:#333,stroke-width:2px
    style D2 fill:#ccf,stroke:#333,stroke-width:2px
    style D3 fill:#ccf,stroke:#333,stroke-width:2px
    style E1 fill:#ccf,stroke:#333,stroke-width:2px
    style E2 fill:#ccf,stroke:#333,stroke-width:2px
    style F1 fill:#ccf,stroke:#333,stroke-width:2px
    style F2 fill:#ccf,stroke:#333,stroke-width:2px
```
    \end{mermaid}
    \caption{Software Overview of pyFAST Library}
    \label{fig:software_overview}
\end{figure}
\texttt{pyFAST} is architected around modularity, with each component engineered to address a specific facet of time series analysis. This modular design philosophy enhances flexibility, maintainability, and extensibility, allowing users to tailor the library to diverse analytical tasks. The library’s core functionalities are organized into five modules, each designed with specific capabilities to ensure a cohesive and versatile framework. The \texttt{data} module is dedicated to data handling and preprocessing, featuring dataset classes tailored for STS, MTM, BDP, and STM scenarios, alongside a suite of scaling methods. This design allows for seamless integration of custom data pipelines.  The \texttt{model} module hosts a diverse collection of time series models, including classical, deep learning (CNNs, RNNs, Transformers), and GNN-based architectures for both UTS and MTS data. The \texttt{base} submodule provides fundamental building blocks for custom model creation.  The \texttt{train} module provides the \texttt{Trainer} class, streamlining model training with functionalities for validation, early stopping, checkpointing, and support for various optimization and learning rate scheduling techniques. The \texttt{metric} module offers a comprehensive suite of evaluation metrics for time series tasks, including specialized metrics for masked data, with the \texttt{Evaluator} class simplifying results reporting.  Lastly, the \texttt{visualize} module provides tools for visualizing time series data and model predictions, aiding in model analysis and interpretation. \texttt{pyFAST}'s modular architecture empowers researchers and practitioners to extend the library, customize modules, combine modules, and reproduce state-of-the-art models.

\texttt{pyFAST} is architected around modularity, with each component engineered to address a specific facet of time series analysis. This modular design philosophy enhances flexibility, maintainability, and extensibility, allowing users to tailor the library to diverse analytical tasks. The library’s core functionalities are organized into five modules, each designed with specific capabilities to ensure a cohesive and versatile framework. The \texttt{data} module is dedicated to data handling and preprocessing, featuring dataset classes tailored for STS, MTM, BDP, and STM scenarios, alongside a suite of scaling methods. This design allows for seamless integration of custom data pipelines.  The \texttt{model} module hosts a diverse collection of time series models, including classical, deep learning (CNNs, RNNs, Transformers), and GNN-based architectures for both UTS and MTS data. The \texttt{base} submodule provides fundamental building blocks for custom model creation.  The \texttt{train} module provides the \texttt{Trainer} class, streamlining model training with functionalities for validation, early stopping, checkpointing, and support for various optimization and learning rate scheduling techniques. The \texttt{metric} module offers a comprehensive suite of evaluation metrics for time series tasks, including specialized metrics for masked data, with the \texttt{Evaluator} class simplifying results reporting.  Lastly, the \texttt{visualize} module provides tools for visualizing time series data and model predictions, aiding in model analysis and interpretation. \texttt{pyFAST}'s modular architecture empowers researchers and practitioners to extend the library, customize modules, combine modules, and reproduce state-of-the-art models.

\section{Comparison with Related Works}

The landscape of time series modeling libraries and frameworks is rich and diverse, with several excellent tools available to researchers and practitioners. TensorFlow Time Series (Abadi et al., 2016) provides a comprehensive and mature toolkit within the TensorFlow ecosystem, with a strong focus on forecasting and anomaly detection.  GluonTS (Alexandrov et al., 2020), developed by Amazon, is a widely adopted library renowned for its extensive suite of probabilistic forecasting models and specialized tools for handling probabilistic time series prediction.  PyTorch Forecasting (Brian et al., 2019) offers a collection of time series-specific layers, models, and training utilities within the PyTorch framework, emphasizing explainable and interpretable forecasting methodologies.  `sktime' (Loning et al., 2019) stands out as a comprehensive Python toolbox encompassing a wide range of time series analysis tasks, including classical models, machine learning models, and sophisticated evaluation tools, with a strong emphasis on seamless integration with the scikit-learn ecosystem.  `tslearn' (Tavenard et al., 2020) specializes in time series-specific machine learning algorithms, offering tools for time series clustering, classification, representation learning, and related tasks.  StatsForecast (Garza et al., 2022) and NeuralForecast (Olivares et al., 2022) are highly optimized libraries specifically engineered for fast and scalable forecasting, providing efficient implementations of both statistical and neural forecasting models, respectively, with a focus on computational performance.

@@

\section{Conclusion}
In summary, \texttt{pyFAST} is presented as a novel, modular, efficient, and comprehensive Python library, poised to significantly benefit both researchers and practitioners in the field of time series analysis. Its core strengths lie in its research-driven flexibility, diverse model library, and ease of use.  Our benchmarking results confirm its practical utility and competitive efficiency against state-of-the-art baselines. We are confident that \texttt{pyFAST} will serve as a valuable, high-performance, and adaptable platform, fostering future innovations in time series research and a wide range of applications. \texttt{pyFAST} is publicly available on GitHub at \href{https://github.com/anonymous-code-for-peer-review/pyFAST}{GitHub}, accompanied by extensive resources to support and expand its community of users and contributors.
